Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-03 documents/ms.tex
adding MAD/MRD table

\newcommand{\columbia}{1}
\newcommand{\oxford}{3}
\newcommand{\sagan}{4}
\newcommand{\uw}{5}
\newcommand{\princeton}{2}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\nkois}{1102}
\newcommand{\nkoimcq}{275}
% \newcommand{\nkoigood}{222}
% \newcommand{\nkoibad}{49}
\newcommand{\kepexample}{5809890}
\newcommand{\kepexampleperiod}{30.5}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{0.60}
\newcommand{\pgramRMS}{0.98}
\newcommand{\mcmcRMS}{0.46}

@@

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarize the results of this MCMC fitting procedure compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
Both the informative and uninformative prior versions of the GP method perform
better than the ACF and periodogram methods, with respective MADs of 0.38 and
0.48, compared to \acfRMS\ (ACF) and \pgramRMS\ (periodogram).
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

@@

We applied the ACF method to the sample of \naigrain\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_acf}.
The $2n$ and $\frac{1}{2}n$ harmonic lines are shown as dashed lines.
The Median Absolute Deviation (MAD) of the ACF-recovered periods is \acfRMS\
in log-space (1.8 days).
The agreement between the injected and recovered rotation periods is good in
general, however some rotation periods are drastically over- and
underestimated.
Of the points clustered close to the $x=y$ line, more fall slightly below it
than above it, \ie\ rotation periods are preferentially underestimated.
This is a feature of the peak position measurement method that is performed on
the ACF.
ACFs of stellar light curves have similar functional forms to the QP kernel
function in equation \ref{eq:QP}: periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left, the
short period end, because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect, however the standard practice is to
simply measure the peak position without taking it into account.
We adopt the standard practice approach here in order to provide a comparison
between our new method and that used largely in the literature.

@@

For each simulated light curve, a LS periodogram
\footnote{LS periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.}
was computed over a grid of 10,000 periods, evenly spaced in frequency,
between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
The MAD for rotation periods measured using the LS periodogram technique is
\pgramRMS\ in log-space (2.7 days).
There is a large number of drastically overestimated periods recovered using
the periodogram.
Excess power at long periods can be generated by long-term trends that are not
related to rotation.
This problem becomes exacerbated when instrumental noise is also present in
light curves.

@@

>>>>

\newcommand{\columbia}{1}
\newcommand{\oxford}{3}
\newcommand{\sagan}{4}
\newcommand{\uw}{5}
\newcommand{\princeton}{2}

\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\nkois}{1102}
\newcommand{\nkoimcq}{275}
\newcommand{\kepexample}{5809890}
\newcommand{\kepexampleperiod}{30.5}
\newcommand{\aigrainexampleperiod}{20.8}

\newcommand{\lnacfMAD}{0.04}
\newcommand{\acfMAD}{0.60}
\newcommand{\percentacfMAD}{3.65}
\newcommand{\lnpgramMAD}{0.04}
\newcommand{\pgramMAD}{0.98}
\newcommand{\percentpgramMAD}{4.32}
\newcommand{\lngpMADnp}{0.02}
\newcommand{\gpMADnp}{0.48}
\newcommand{\percentgpMADnp}{2.45}
\newcommand{\lngpMAD}{0.02}
\newcommand{\gpMAD}{0.38}
\newcommand{\percentgpMAD}{1.85}

@@

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarize the results of this MCMC fitting procedure compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
In order to assess the performance of the GP and other period recovery
methods, we compute Median Absolute Deviations (MADs) of the results, relative
to the input periods.
We also compute the Median Relative Deviation (MRD), as a percentage.
These metrics are presented for the three different methods tested in this
paper in table \ref{tab:MADs}.
The informative and uninformative prior versions of the GP method have MRDs of
\percentgpMAD\ \% and \percentgpMADnp \% respectively.
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

@@

We applied the ACF method to the sample of \naigrain\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_acf}.
The $2n$ and $\frac{1}{2}n$ harmonic lines are shown as dashed lines.
The MRD of the ACF-recovered periods is \percentacfMAD\ \% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
The agreement between the injected and recovered rotation periods is good in
general, however some rotation periods are drastically over- and
underestimated.
Of the points clustered close to the $x=y$ line, more fall slightly below it
than above it, \ie\ rotation periods are preferentially underestimated.
This is a feature of the peak position measurement method that is performed on
the ACF.
ACFs of stellar light curves have similar functional forms to the QP kernel
function in equation \ref{eq:QP}: periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left, the
short period end, because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect, however the standard practice is to
simply measure the peak position without taking it into account.
We adopt the standard practice approach here in order to provide a comparison
between our new method and that used largely in the literature.

@@

For each simulated light curve, a LS periodogram
\footnote{LS periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.}
was computed over a grid of 10,000 periods, evenly spaced in frequency,
between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
The MRD of the pgram-recovered periods is \percentpgramMAD\ \% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
There is a large number of drastically overestimated periods recovered using
the periodogram.
Excess power at long periods can be generated by long-term trends that are not
related to rotation.
This problem becomes exacerbated when instrumental noise is also present in
light curves.

@@

\begin{table*}
\begin{center}
    \caption{Median absolute (MAD) and median relative (MRD) deviations for
    the ACF, LS periodogram and GP period recovery methods.}
\begin{tabular}{lccc}
Method & MAD & MRD \\
    \hline
    ACF & \acfMAD\ days & \percentacfMAD\ \% \\
    LS periodogram & \pgramMAD\ days & \percentpgramMAD\ \% \\
    GP (no prior) & \gpMADnp\ days & \percentgpMADnp\ \% \\
    GP (acf prior) & \gpMAD\ days & \percentgpMAD\ \% \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:MADs}
